Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites

### 2021-03-25 doc/SA-Hermaphrodites-wDemography.tex
Minor edits to the main text

\documentclass[11pt]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}

@@

\begin{document}
\title{Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites}
\author{Colin Olito$^{1,\ast}$ \\ 
Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

@@

%\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section*{Abstract -- currently 215/200 word max.}
%%%%%%%%%%%%%%%%%%%%
Genetic trade-offs between male and female fitness impose fascinating genetic constraints on adaptation. On one hand they can maintain genetic variation for a variety of life-history and gender-related traits, while on the other hand they cause unavoidable maladaptation as individuals are prevented from reaching their sex-specific fitness optima. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of sexually antagonistic (SA) selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection {\itshape and} the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages. Finally, we gave an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

\newpage{}
\section*{ALTERNATIVE Abstract -- currently 200/200 word max.}
% Alternative opening sentence to the abstract:sentence
Now classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection and the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant Mimulus guttatus. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with M. guttatus appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

In order to persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, changing environmental conditions \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

@@

Sexually antagonistic selection (abbreviated SA hereafter) arising from genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and also the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, SA selection can arise when beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}. Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}.

The evolutionary dynamics of sexually antagonistic alleles in hermaphrodites differ from those in dioecious populations because individuals may reproduce through a combination of self- and outcross fertilization \citep{Goodwillie2005, Igic2006, JarneAuld2006}. For example, self-fertilization is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019}. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals. To take the population dynamical consequences of polymorphism into account, a model is required that links changes in genotype distributions to population dynamics. For example, \cite{harts2014demography} developed an individual-based modelling approach that tracks population dynamics and the spread of both SA genes and dispersal genes in a species with non-overlapping generations. \cite{deVriesCaswell2019b} introduced a mendelian matrix model with overlapping generations, intralocus sexual antagonism, and population dynamics but did not perform analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

@@

\noindent The matrix $\mbb{F}^{\prime}$ is composed of genotype-specific fertility matrices, and operates on the the vector of genotype frequencies to give their relative contribution of each genotype to the gamete pool. The matrix $\mbf{W}^{\prime}$ converts these contributions to allele numbers. Normalizing the resulting vector gives the allele frequencies in the male gamete pool:
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
			\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime} \|} = 
				\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)\|}
\end{equation}
\end{linenomath*}
The key difference between selfing and outcrossing reproduction can be seen in the parent-offspring maps, which reflect the joint effects of meiosis and mating on the distribution of offspring genotypes.
% \textcolor{red}{
% I'm doubting whether to move the next paragraph to the appendices as well.  Equation (\ref{eq:maleGametePool}) is all we strictly need. If we keep the parent-offspring maps, we can explain in the next section how the matrices $\mcal{F}^S$ and $\mcal{F}^X$ have the same structure. Would that be helpful? Or would it be easier to just explain in the next section what the structure of $\mcal{F}^S$ and $\mcal{F}^X$ is? }

The parent-offspring map for outcrossing is a function of the allele frequencies in the male gamete pool, and is given by 
\begin{linenomath*}
\begin{equation} \label{eq:HX}
	\mbf{H}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} & \frac{1}{2} q^{\prime}_{A} & 0 \\
				q^{\prime}_{a} & \frac{1}{2} & q^{\prime}_{A}  \\
				0 & \frac{1}{2} q^{\prime}_{a} & q^{\prime}_{a} \\
			\end{array} \right)
\end{equation}
\end{linenomath*}
\noindent From left to right, the columns of matrix $\mbf{H}^X$ give the genotype distribution of outcrossed offspring produced by a maternal parent of each genotype ($AA$, $Aa$, and $aa$ respectively).

Reproduction via self-fertilization differs from outcrossing. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively. As described in the next section, these parent-offspring maps influence the projection matrix by altering the fertilitiy matrices for individuals produced by selfing and outcrossing (see Eq(\ref{eq:BlkFS}) and Eq(\ref{eq:BlkFX}) below).

@@

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of selfed and outcrossed offspring ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrix for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}). Further details of the derivations of $\mcal{U}^S$, $\mcal{U}^X$ ,  $\mcal{F}^S$, and $\mcal{F}^X$ are provided in \hl{Appendix A.x}.

@@

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. To keep the model analytically tractable, we assume our hypothetical species has two life stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect.

@@

Iterating the projection matrix, Eq(\ref{eq:Atilde}) with the above demographic matrices, given an initial population state vector, Eq(\ref{eq:PopStateVec}), allows numerical simulation of the eco-evolutionary dynamics for selection operating on any of the sex-function, or stage-specific demographic parameters. As we outline below, we use numerical techniques together with mathematical analyses to study the conditions for the maintenance of SA polymorphisms, and the demographic fate of the populations (i.e., positive growth, or extinction).

@@

Diverse eco-evolutionary outcomes are possible in the model, including fixation of either allele, balanced polymorphism, population growth or extinction, and even evolutionary rescue and suicide (e.g., see \citealt{CaswellEtAl2018,deVriesCaswell2019a,deVriesCaswell2019b}). We focus on identifying parameter conditions where two criteria are satisfied: ({\itshape i}) SA polymorphism is maintained under balancing selection and ({\itshape ii}) the intrinsic population growth rate at equilibrium is positive; a situation that we refer to as a 'demographically viable SA polymorphism'.

@@

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. In our model, the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see \citealt{Caswell2001}), and the intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible \citep{de2020matrix}.

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we also assume equal dominance in all analyses (i.e., $h_f = h_m = h$). For a given parameter set, we numerically evaluated the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrated inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

@@

\textcolor{red}{This needs some polishing, but I though you may want to embellish.}. \cite{grieshop2017male} found that genotypes with sexually antagonstic alleles that are male beneficial experience higher levels of inbreeding depression than genotypes with female beneficial SA alleles. This effect can easily be included in our model by making the inbreeding depression parameters, $\delta$ and $\delta_i$, a function of the genotype of the individual. Such an effect would further reduce the demographically viable polymorphic parameter space, and increase the bias in viable SA polymorphisms towards alleles with weaker selection in females.

>>>>

\documentclass[11pt,draft]{article}
% Preamble
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\bibliographystyle{amnatnat.bst}
\setlength{\bibsep}{0.0pt}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage[left]{lineno}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[utf8]{inputenc}
\usepackage{color,soul}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{pdflscape}
\usepackage{nameref}
%\usepackage[title,titletoc,toc]{appendix}
\usepackage[colorlinks=true, allcolors=black]{hyperref}

@@

\begin{document}
\title{Evolutionary demography and the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites}

\author{Colin Olito$^{1,\ast}$ \\ 
Charlotte de Vries$^{2}$}
\date{\today}
\maketitle

@@

%\tableofcontents
%%%%%%%%%%%%%%%%%%%%
\newpage{}
\section{Title options? \title{Population viability affects/impacts the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites}}

 \section{Demographic viability of antagonistic alleles impacts conditions for the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites}

 \section{Conditions for the maintenance of sexually antagonistic polymorphism in outcrossers and simultaneous hermaphrodites impacted by considering population viability}

\section*{Abstract -- currently 215/200 word max.}
%%%%%%%%%%%%%%%%%%%%
Genetic trade-offs between male and female fitness impose fascinating genetic constraints on adaptation. On one hand they can maintain genetic variation for a variety of life-history and gender-related traits, while on the other hand they cause unavoidable maladaptation as individuals are prevented from reaching their sex-specific fitness optima. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of sexually antagonistic (SA) selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained {\itshape and} the population intrinsic growth rate remains positive. We also analyze the effects of inbreeding depression manifesting in different life-history stages. Finally, we show an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant {\itshape Mimulus guttatus}. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with {\itshape M.~guttatus} appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

\newpage{}
\section*{ALTERNATIVE Abstract -- currently 200/200 word max.\textcolor{green}{ I like this abstract best}}

% Alternative opening sentence to the abstract:sentence
Classic population genetics theory has been fundamental to understanding the evolution of sex-differences and the maintenance of sexually antagonistic (SA) genetic variation, but these models have rarely considered the demographic consequences of intralocus sexual antagonism. In this paper we develop a stage-structured mendelian matrix model and jointly analyze the evolutionary and demographic consequences of SA selection in obligately outcrossing (equivalent to dioecious) and partially selfing hermaphrodite populations. We focus on identifying the parameter conditions under which SA polymorphisms are maintained by balancing selection \textit{and} the population intrinsic growth rate remains positive. In addition, we analyze the effects of inbreeding depression manifesting in different life-history stages, and give an illustrative example of the potential for SA polymorphism in real populations using empirically estimated demographic rates for the hermaphroditic flowering plant Mimulus guttatus. Our results indicate that when population growth rates are near one, large swathes of polymorphic parameter space become demographically inviable, and that inbreeding depression poses a significant challenge to the maintenance of SA polymorphism in partially selfing populations. Despite these demographic challenges, our example with M. guttatus appears to show that demographic rates observed in some real populations are capable of sustaining large regions of viable SA polymorphic space.

@@

To persist in the long term, a population must be able to adapt to its environment. Yet, the process of adaptation can be impeded by a variety of genetic and environmental factors, such as deleterious mutations \citep{Haldane1957}, changing environmental conditions \citep{Maynard-Smith1976, LandeShannon1996,OrrUnckless2008}, maladaptive gene-flow \citep{KirkpatrickBarton1997, BolnickNosil2007}, and genetic constraints \citep{ConnallonHall2018, MatthewsConnallon2019}. Genetic constraints on adaptation can arise as a consequence of conflicting selection and gene-flow between different classes of individuals, or more generally, when there are genetic trade-offs between fitness components \citep{CharlesworthHughes2000, ConnallonHall2018}.

@@

Sexually antagonistic selection (abbreviated SA hereafter) arising from genetic trade-offs between male and female fitness, is a common feature of sexually reproducing populations, and is thought to contribute to the maintenance of genetic variation and the evolution of sexual dimorphism and gender-related traits \citep{Lande1980, Rice1992, Charlesworth1999, RiceChippindale2001, BondurianskyChenoweth2009,Olito2019}. When such genetic trade-offs occur, SA selection can arise when beneficial alleles for one sex are deleterious when expressed in the other \citep{Kidwell1977, Rice1992, ConnallonClark2012}.

Alleles with opposing fitness effects through male and female sex functions can cause analogous genetic constraints on fitness in hermaphrodite populations, where both maternal and paternal reproductive success contribute jointly to each individuals' overall fitness \citep{LloydWebb1986, WebbLloyd1986, Abbott2011, JordanConnallon2014}. Self-fertilization in hermaphrodites is predicted to reduce the total parameter space where balanced SA polymorphisms can be maintained, while simultaneously creating a bias in selection through the female sex function (\citealt{JordanConnallon2014}; but see \citealt{Tazzyman2015}). Yet, other factors, such as genetic linkage to other SA loci or a sex-determining region \citep{Otto2011, JordanCharlesworth2012, Olito2017, Olito2019}, or spatial heterogeneity and complexity of the life-cycle can expand the parameter space for SA polymorphism \citep{Olito-etal-2018,ConnallonSharmaOlito2019}. Overall, both theoretical predictions and current empirical data suggest there is ample scope for SA trade-offs and the maintenance of SA polymorphisms in both dioecious and hermaphrodite populations \citep{Abbott2011, WangBarrett2020}, although identifying specific SA loci from genome sequence data remains challenging \citep{RuzickaESEB2020}.

Most population genetic models, including models of the maintenance of SA polymorphisms, keep track of the relative fitnesses of genotypes in populations of constant size, and therefore do not consider the population dynamical consequences of evolutionary change. In fact, these genotypic fitnesses emerge from myriad processes acting throughout the life-history of individuals (e.g., \cite{johnston2013life},\cite{merot2020balancing}). To take the population dynamical consequences of polymorphism into account, a model is required that links changes in genotype distributions to population dynamics. The potential demographic costs of sexual antagonism were pointed out by \cite{kokko2003sexy}, but few papers since then have taken this into account (with the notable exception of \cite{harts2014demography}). \cite{deVriesCaswell2019b} introduced a mendelian matrix model with intralocus sexual antagonism, and population dynamics but did not perform an analysis of the consequences of demographic viability for the scope of sexual antagonism to maintain polymorphism.

@@

\noindent The matrix $\mbb{F}^{\prime}$ is composed of genotype-specific fertility matrices, and operates on the the vector of genotype frequencies to give their relative contribution of each genotype to the gamete pool. The matrix $\mbf{W}^{\prime}$ converts these contributions to allele numbers. Normalizing the resulting vector gives the allele frequencies in the male gamete pool:
\begin{linenomath*}
\begin{equation} \label{eq:maleGametePool}
	\left(
		\begin{array}{c}
			q^{\prime}_{A} \\
			q^{\prime}_{a} \\
		\end{array} \right) = 
			\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime}}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} \mbf{p}^{\prime} \|} = 
				\frac{\mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)}{\| \mbf{W}^{\prime} \mbb{F}^{\prime} (\mbf{n}^X + \mbf{n}^S)\|}.
\end{equation}
\end{linenomath*}
The key difference between selfing and outcrossing reproduction can be seen in the parent-offspring maps, which reflect the joint effects of meiosis and mating on the distribution of offspring genotypes.
% \textcolor{red}{
% I'm doubting whether to move the next paragraph to the appendices as well.  Equation (\ref{eq:maleGametePool}) is all we strictly need. If we keep the parent-offspring maps, we can explain in the next section how the matrices $\mcal{F}^S$ and $\mcal{F}^X$ have the same structure. Would that be helpful? Or would it be easier to just explain in the next section what the structure of $\mcal{F}^S$ and $\mcal{F}^X$ is? }

The parent-offspring map for outcrossing is a function of the allele frequencies in the male gamete pool, and is given by 
\begin{linenomath*}
\begin{equation} \label{eq:HX}
	\mbf{H}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} & \frac{1}{2} q^{\prime}_{A} & 0 \\
				q^{\prime}_{a} & \frac{1}{2} & q^{\prime}_{A}  \\
				0 & \frac{1}{2} q^{\prime}_{a} & q^{\prime}_{a} \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent From left to right, the columns of matrix $\mbf{H}^X$ give the genotype distribution of outcrossed offspring produced by a maternal parent of each genotype ($AA$, $Aa$, and $aa$ respectively).

The parent-offpsring map for reproduction via self-fertilization differs from the outcrossing parent-offspring map. As in previous models of partial selfing, we assume that individuals produce enough male gametes to easily self-fertilize all of their ovules, and that self-fertilization involves little or no selection from external factors relative to outcrossing (e.g., \citealt{Charlesworth1978a,JordanConnallon2014,Olito2017}, but see \citealt{Tazzyman2015}). Under these assumptions, the genotype distributions of selfed offspring are determined entirely by the parental genotype and the probabilities of segregation and fertilization during and after meiosis: 
\begin{linenomath*}
\begin{equation} \label{eq:HS}
	\mbf{H}^S(\tilde{\mbf{n}}) = 
			\left(
			\begin{array}{ccc}
				1 & 1/4 & 0 \\
				0 & 1/2 & 0 \\
				0 & 1/4 & 1 \\
			\end{array} \right).
\end{equation}
\end{linenomath*}
\noindent The columns of $\mbf{H}^S(\tilde{\mbf{n}})$ again give the selfed offspring genotype distributions for parental genotypes of $AA$, $Aa$, and $aa$, respectively. As described in the next section, these parent-offspring maps influence the projection matrix by altering the fertilitiy matrices for individuals produced by selfing and outcrossing (see Eq(\ref{eq:BlkFS}) and Eq(\ref{eq:BlkFX}) below).

@@

The blocks of the component matrices in Eq(\ref{eq:Atilde}) correspond to production of offspring by self-fertilization and outcrossing ($\mcal{F}^S$ and $\mcal{F}^X$ in $\tilde{\mbf{F}}$), and survival of extant individuals produced by selfing or outcrossing ($\mcal{U}^S$ and $\mcal{U}^X$ in $\tilde{\mbf{U}}$). The survival matrix for individuals produced through selfing and outcrossing are,
\begin{linenomath*}
\begin{eqnarray} 
	\mcal{U}^S  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{S}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{S}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{S}_{aa} \\
			\end{array} \right),\label{eq:BlkUS}\\
				\mcal{U}^X  = 
		\left(
			\begin{array}{ccc}
				\mbf{U}^{X}_{AA} & \mbf{0} & \mbf{0} \\
				\mbf{0} & \mbf{U}^{X}_{Aa} & \mbf{0} \\
				\mbf{0} & \mbf{0} & \mbf{U}^{X}_{aa} \\
			\end{array} \right).\label{eq:BlkUX}
\end{eqnarray}
\end{linenomath*}
Since individuals cannot change their genotype once they are born, the survival matrices are block diagonal.  Similarly, we construct fertility matrices for individuals produced through selfing and outcrossing, 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFS}
	\mcal{F}^S = 
			\left(
			\begin{array}{ccc}
				\mbf{F}_{AA} & \frac{1}{4} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{2} \mbf{F}_{Aa} & 0 \\
				0 & \frac{1}{4} \mbf{F}_{Aa} & \mbf{F}_{aa}\\
			\end{array} \right), 
\end{equation}
\end{linenomath*}
\noindent and 
\begin{linenomath*}
\begin{equation} \label{eq:BlkFX}
	\mcal{F}^X = 
			\left(
			\begin{array}{ccc}
				q^{\prime}_{A} \mbf{F}_{AA} & \frac{1}{2} q^{\prime}_{A} \mbf{F}_{Aa} & 0 \\ 
				q^{\prime}_{a} \mbf{F}_{AA} & \frac{1}{2} \mbf{F}_{Aa} & q^{\prime}_{A} \mbf{F}_{aa}  \\ 
				0 & \frac{1}{2} q^{\prime}_{a} \mbf{F}_{Aa} & q^{\prime}_{a} \mbf{F}_{aa}  \\
			\end{array} \right),
\end{equation}
\end{linenomath*}
\noindent where $q^{\prime}_A$ and $q^{\prime}_a$ are given by Eq(\ref{eq:maleGametePool}). Further details of the derivations of $\mcal{U}^S$, $\mcal{U}^X$ ,  $\mcal{F}^S$, and $\mcal{F}^X$ are provided in \hl{Appendix A.x}.

@@

We now construct and analyze a genotype $\times$ stage-classified model for a hypothetical species with intralocus sexual conflict via the two sex functions. For the sake of simplicity, we assume our hypothetical species has a basic life cycle with only two life stages: juveniles and adults (i.e., $\omega = 2$), and that only adults are reproductively active. Suppose that there is a genetic trade-off between the sex-functions at a single diallelic locus such that allele $A$ is beneficial for female fertility but detrimental for male reproductive success (e.g., pollen production), and that allele $a$ has the reverse effect.

@@

Iterating the projection matrix, Eq(\ref{eq:Atilde}) with the above demographic matrices, given an initial population state vector, allows numerical simulation of the eco-evolutionary dynamics for selection operating on any of the sex-function, or stage-specific demographic parameters. As we outline below, we use numerical techniques together with mathematical analyses to study the conditions for the maintenance of SA polymorphisms, and the demographic fate of the populations (i.e., positive growth, or extinction).

@@

Diverse eco-evolutionary outcomes are possible in the model, including fixation of either allele, balanced polymorphism, population growth or extinction, and even evolutionary rescue and suicide (e.g., see \citealt{deVriesCaswell2019a,deVriesCaswell2019b}). We focus on identifying parameter conditions where two criteria are satisfied: ({\itshape i}) SA polymorphism is maintained under balancing selection and ({\itshape ii}) the intrinsic population growth rate at equilibrium is positive; a situation that we refer to as a 'demographically viable SA polymorphism'.

@@

We used numerical simulation to determine whether a protected SA polymorphism was also demographically viable. Specifically, for each boundary equilibrium we introduced the rare allele at low initial frequency, and iterated Eq(\ref{eq:Atilde}) until the population had reached demographic and genotypic equilibrium. In our model with density independent demographic rates, the population state vector will grow or shrink exponentially after converging to stable population structure and genotypic frequencies (see chapter 17 in \citealt{Caswell2001}), and the intrinsic population growth rate after convergence, $\lambda$, can be calculated as $\tilde{\mbf{n}}(t)/\tilde{\mbf{n}}(t-1)$. We note, however, that if the ecological component of the model is non-linear, more exotic dynamics are possible \citep{de2020matrix}.

% Unfortunately, closed-form solutions for the invasion conditions for each allele and the equilibrium frequencies under different selection scenarios are analytically intractable except for restricted conditions (see \hl{Online Appendix X}). We therefore used numerical techniques to identify invasion and extinction thresholds, which could then be used to define regions of demographically viable polymorphism across $s_f \times s_m$ parameter space. 
Because single-locus selection coefficients are generally weak (i.e., $s < 0.1$; e.g., \citealt{Eyre-WalkerKeightly2007}) and strongly skewed, we limit our analyses to coefficients within $0 < s_f,s_m \leq 0.15$, unless stated otherwise. For simplicity, we also assume equal dominance in all analyses (i.e., $h_f = h_m = h$). For a given parameter set, we numerically evaluate the largest eigenvalue of the Jacobian evaluated at both boundaries, and the population growth rate of the equilibrium obtained after perturbation of the boundary, $\lambda$. To determine the viability threshold (where the population growth rate equals one) and the boundary stability thresholds (where the largest eigenvalue of the Jacobian crosses one), we start at the outer boundaries of selection parameter space, and titrate inward. The proportions of total $s_f \times s_m$ parameter space where SA polymorphism and/or extinction occurred were calculated by numerically integrating the appropriate regions defined by the invasion and extinction thresholds

@@

\cite{grieshop2017male} found that genotypes with sexually antagonstic alleles that are male beneficial experience higher levels of inbreeding depression than genotypes with female beneficial SA alleles. This effect can easily be included in our model by making the inbreeding depression parameters, $\delta$ and $\delta_i$, a function of the genotype of the individual. Such an effect would further reduce the demographically viable polymorphic parameter space, and increase the bias in viable SA polymorphisms towards alleles with weaker selection in females.
